---
title: 'Winds of change: Analyzing 70 years of Tornado data in the US'
author: "Edward Harvey"
date: "`r Sys.Date()`"
output:
  slidy_presentation:
    theme: "default"
    keep_tex: true
    #incremental: true
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r, include=FALSE}
# Load required packages
library(tidyverse)
library(plotly)
library(shiny)
library(lubridate)
library(knitr)
library(kableExtra)
library(RColorBrewer)


setwd("/Users/faculty/Documents/UMASS/STAT 697V/Final Presentation")

# Load the tornado dataset
df <- read.csv("us_tornado_dataset_1950_2021.csv")

df$date <- ymd(df$date)

state_avg_temp <- readRDS("state_avg_temp.rds")
states <- c("AK", "AL", "AR", "AZ", "CA", "CO", "CT", "DE", "FL", "GA", "IA", "ID", "IL", "IN", "KS", "KY", "LA", "MA", "MD", "ME", "MI", "MN", "MO", "MS", "MT", "NC", "ND", "NE", "NH", "NJ", "NM", "NV", "NY", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", "VA", "VT", "WA", "WI", "WV", "WY")




state_avg_temp_long <- pivot_longer(state_avg_temp, cols = c(states), values_to = "avg_temp", names_to = "st")


df <- left_join(x = df, y = state_avg_temp_long, by = c("yr", "st"))

df$avg_temp <- as.numeric(df$avg_temp)

#group by decade
df$decade <- paste0(substr(df$yr,1,3),0, "s")

saveRDS(df, "df_cleaned.rds")



```

# Research questions:

Climate scientists have warned that climate change could cause more severe storms. In the US, tornados cause massive destruction each year. 

- Have measures of severity gotten worse over time?
- Do rising temperatures predict more severe storms?
- Has the geographic distribution of storms changed?

# The dataset

70 years of tornado records (1950 to 2021) across the US including variables about date, location, severity, and casualties. Accessed from [kaggle.com](https://www.kaggle.com/datasets/danbraswell/us-tornado-dataset-1950-2021).

```{r}
kable(head(df)) %>% kable_styling(font_size = 18, full_width = TRUE)
```


```{r, include=FALSE}


df_by_decade <- df %>% filter(mag >-1) %>% group_by(decade, mag) %>% summarise(mag = mean(mag),
                                                      fat = sum(fat),
                                                      inj = sum(inj),
                                                      wid = mean(wid),
                                                      len = mean(len),
                                                      count = n()) %>% ungroup()



saveRDS(df_by_decade, "df_by_decade.rds")
```

# Exploring tornado frequency

It appears that tornadoes have become a more frequent occurrence in recent decades. (Note: the 2020s are not yet complete).

```{r, out.width="90%"}
#frequency plot
freq_by_decade <- plot_ly(data =df_by_decade) %>% 
  add_trace(x=~decade,
            y=~count,
            name = "Total",
            type = "bar")

freq_by_decade_split <- plot_ly(data =df_by_decade) %>% 
  add_trace(x=~decade,
            y=~count,
            type = "bar",
            split = ~mag)


subplot(freq_by_decade, freq_by_decade_split, nrows = 2)

```


# Fatalities have declined

```{r, out.width="90%"}
fat_by_decade <- plot_ly(data = df_by_decade) %>% 
  add_trace(x = ~decade,
            y= ~fat,
            name = "Total",
            type = "bar")

fat_by_decade_split <- plot_ly(data = df_by_decade) %>% 
  add_trace(x = ~decade,
            y= ~fat,
            type = "bar",
            split = ~mag)

subplot(fat_by_decade, fat_by_decade_split, nrows = 2)

```

# Injuries have also declined

```{r, out.width="90%"}
inj_by_decade <- plot_ly(data = df_by_decade) %>% 
  add_trace(x = ~decade,
            y= ~inj,
            name = "Total",
            type = "bar")

inj_by_decade_split <- plot_ly(data = df_by_decade) %>% 
  add_trace(x = ~decade,
            y= ~inj,
            type = "bar",
            split = ~mag)


subplot(inj_by_decade, inj_by_decade_split, nrows=2)
```

# Storm width is increasing, track length is not

```{r, out.width="90%"}
wid_over_time <- plot_ly(data = df_by_decade) %>% 
  add_trace(x = ~decade,
            y = ~wid,
            name = "width",
            type = "bar")

len_over_time <- plot_ly(data = df_by_decade) %>% 
  add_trace(x = ~decade,
            y = ~len,
            name = "length",
            type = "bar")

subplot(wid_over_time, len_over_time, nrows = 2)


```

# Force category across the decades

A chi square test for homogeneity gives significant results, but heatmap visualization casts doubt on whether we should reject the null hypothesis. 


```{r,echo=FALSE, warning=FALSE, message=FALSE}
shinyAppFile("chi2app.R")
```


# Average temperatures and tornado counts are rising
```{r, warning = FALSE, message=FALSE, out.width="90%"}

US_by_year <- df %>% filter(mag >-1) %>% group_by(yr) %>% summarise(mag = mean(mag),
                                                      fat = sum(fat),
                                                      inj = sum(inj),
                                                      wid = mean(wid),
                                                      len = mean(len),
                                                      avg_temp = mean(avg_temp, na.rm = TRUE),
                                                      count = n()) %>% ungroup()


#temperature over time
temp_model <- lm(data = US_by_year, avg_temp ~ yr)
pred_temp <- predict.lm(temp_model, newdata = US_by_year)



temp_over_time <- plot_ly(data = US_by_year) %>% 
  add_markers(x = ~yr, y = ~avg_temp, type = "scatter", name = "average temperature") %>% 
  add_trace(x = ~yr, y = ~pred_temp, mode = "lines", showlegend = FALSE) 


count_over_time <- plot_ly(data = US_by_year) %>% 
  add_markers(x = ~yr, y = ~count, type = "scatter", name = "count")

subplot(temp_over_time, count_over_time, nrows = 2)
```


# No clear relationship between temperature and tornado frequency
```{r, out.width="90%"}


plot_ly(data = US_by_year) %>% 
  add_markers(x = ~avg_temp, y = ~count, type = "scatter", name = "count") %>% 
  add_paths(x = ~avg_temp, y = ~count)
```


# Width increases with temperature (yearly)
```{r, warning = FALSE, message=FALSE, out.width="90%"}
#width vs temperature
wid_model <- lm(data = US_by_year, wid ~ avg_temp)
pred_wid <- predict.lm(wid_model, newdata = US_by_year)

wid_v_temp <- plot_ly(data = US_by_year) %>%
  add_markers(x = ~avg_temp, y=~wid, type = "scatter", name = "width") %>%
  add_trace(x = ~avg_temp, y = ~pred_wid, mode = "lines", showlegend = FALSE)

#length vs temperature
len_model <- lm(data = US_by_year, len ~ avg_temp)
pred_len <- predict.lm(len_model, newdata = US_by_year)

len_v_temp <- plot_ly(data = US_by_year) %>%
  add_markers(x = ~avg_temp, y=~len, type = "scatter", name = "length") %>%
  add_trace(x = ~avg_temp, y = ~pred_len, mode = "lines", showlegend = FALSE)

#magnitude vs temperature
mag_model <- lm(data = US_by_year, mag ~ avg_temp)
pred_mag <- predict.lm(mag_model, newdata = US_by_year)

mag_v_temp <- plot_ly(data = US_by_year) %>%
  add_markers(x = ~avg_temp, y=~mag, type = "scatter", name = "magnitude") %>%
  add_trace(x = ~avg_temp, y = ~pred_mag, mode = "lines", showlegend = FALSE)

subplot(wid_v_temp, len_v_temp, mag_v_temp, nrows = 3, shareX = TRUE, titleX = TRUE)

```


# Geographic distribution of storms appears consistent

```{r,echo=FALSE, warning=FALSE, message=FALSE}
shinyAppFile("map_app_V2.R")
```


# Summary

## **Conclusions**

- Initial findings suggest tornadoes are becoming more frequent and wider over time
- Fatalities, injuries, and magnitude have not increased
- While temperatures have increased, it's not clear that temperature is correlated with severity
- No obvious change in geographic distribution
- Even highly local WX reports do not give indications of tornado activity

## **Considerations**
- Precision of measurement and technology are likely confounding factors
- More granular weather data is needed
- In the age of accurate forecasting, property damage may be a more relevant measure of severity than casualties




